Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-01 ind_files/01_Introduction.tex
Add ito formula in background
Significantly shorten and smoothen intro

More than 150 years have passed since Ernst Haeckel first coined the term ‘ecology’ in 1866. Today, ecology is incredibly interdisciplinary, borrowing techniques and ideas from diverse fields such as computer science, statistics, economics, dynamical systems, physics, and information theory. With this development has come an array of tools intended to help the theoretician use abstract entities which only exist on paper to capture the essence of the real biological systems we see around us. Consequently, there has also been a boom in models that seek to explain natural phenomena using these tools. Despite the apparent overabundance of such seemingly disparate models in the literature, theoretical ecologists often think in terms of a few very general organizing principles which can be adapted in various ways to model specific biological systems \citep{vellend_theory_2016}. There is thus value to formulating general, idealized models in abstract terms to underscore these organizing principles and better illustrate the fundamental processes that are required to capture the `essence' of an ecological pattern. Such abstractions and idealizations are vital to model building and are useful in gaining insights into complex phenomena such as those that abound in ecology. However, it is important to remember that abstractions, are, ultimately, unrealistic by construction. As such, it is important to investigate whether predictions of general abstract theories are robust to failure of some of the more unrealistic assumptions that they make. Some of the central questions of ecology boil down to the question of investigating the phenomenon of multiple interacting ‘eco-variants’ (ecologically relevant phenotypic variants: Some examples in section \ref{synthesis}) coexisting in sympatry. In section \ref{idealization}, I reiterate some old arguments about the utility and pitfalls of general idealized models in ecology and evolution. In section \ref{synthesis}, I highlight diverse ecologically important contexts in which the coexistence of multiple discrete `eco-variants' is relevant. One general class of models, called `adaptive diversification', has gained immense popularity in trying to answer such questions of coexisting phenotypic variants. Chapter \ref{chap_AD} introduces this research program. I summarize three distinct modelling approaches to adaptive diversification in section \ref{AD} that differ in their simplifying assumptions and present some important general predictions of these models in sections \ref{high_dim} and \ref{dem_stoch}. I identify a problematic idealization within these existing models, and outline my goal of addressing this shortcoming for my thesis in section \ref{goals}. Chapter \ref{chap_BD} develops stochastic models of eco-evolutionary population dynamics from first principles, and I show that various known models from evolutionary game theory, population genetics, quantitative dynamics, and adaptive dynamics arise as special cases in the deterministic limit. This section also highlights the connections between the three modelling paradigms of adaptive diversification mentioned in section \ref{AD} by explicitly deriving them. This modelling exercise results in a useful organizational framework for general population models in ecology and evolution and shows how tools from statistical physics and mathematics can be useful in analyzing the models. Some approximations derived in this chapter are used in Chapter \ref{chap_examples} to analytically investigate adaptive diversification in finite populations.

@@

\section{Coexistence of eco-variants in diverse contexts}\label{synthesis}
The phenomenon of coexistence of multiple phenotypically distinct, ecologically relevant variants of an organism in sympatry occurs in several different contexts in the natural world. This is best illustrated through examples, which I provide below.
\subsection{Alternative Reproductive Tactics}
Alternative reproductive tactics (ARTs) are discrete polymorphisms that occur in sympatry in one or both sexes of a species in response to intrasexual competition for mates. More precisely, two or more traits which are expressed in individuals of a given sex of a given species can be said to be ARTs if \citep{oliveira_alternative_2008}:
\begin{enumerate}[label=(\alph*)]
    \item They are \emph{alternative}, in the sense that a given individual can only express one of the behaviors at any given point in time.
    \item They are \emph{reproductive}, in the sense that the traits are directly relevant to the process of obtaining a mate and are intended to mitigate conspecific intrasexual competition.
    \item They are \emph{tactics}, in the sense of serving a well-defined adaptive function and thus having fitness consequences. Usually, different ARTs have different associated fitness effects \emph{ceteris paribus}.
\end{enumerate}
ARTs are widespread in the animal kingdom, and may be seen either in morphological traits, or as more complex behavioral polymorphisms (see \citep{oliveira_alternative_2008} for an overview). Most (if not all) species which exhibit ARTs only show a small number of polymorphisms (usually 2-3, 5 in a handful of cases, and very rarely 7-8) - \hl{I can maybe make a table for this: Somehow wasn't able to find one in the literature}. ARTs have obvious consequences for reproductive life history, and can strongly influence the ecology and evolutionary trajectory of organisms.

\subsection{Trophic resource polymorphisms}

Trophic resource polymorphism is defined as ``the occurrence of discrete intraspecific morphs showing differential niche use, usually through differences in feeding biology and habitat use" \citep{skulason_resource_1995}. Species which exhibit trophic resource polymorphisms often show marked, discontinuous morphological variations adapted for specific niche usage. For example, arctic charr and african cichlid fishes both often exhibit discrete trophic polymorphisms in which different morphs (of the same species) are specialized for feeding at different strata of the lakes they live in \citep{recknagel_ecosystem_2017}. Such polymorphisms also generally occur in relatively small numbers in sympatry (see Table 1 in \citep{smith_evolutionary_1996}). They are relatively widespread in vertebrates, and are thought to have important evolutionary consequences by acting as the `substrate' for adaptive radiations \citep{smith_evolutionary_1996}.

\subsection{Mating types in isogamous species}
Isogamous organisms are sexual, but do not have gametes which can be classified into `male' and `female' based on size. Gametes of such species can, however, be divided into distinct `mating types' according to chemical self-incompatibility. There is substantial variation in the number of mating types present in a given species: While most organisms have only two mating types, this is by no means the rule \citep{phadke_rapid_2009,constable_rate_2018}, with a handful of organisms having 7-10 mating types, and one species of fungus (\textit{Schizophyllum commune}) exhibiting over 23,000 mating types. The study of mating types in isogamous species is ecologically and evolutionarily relevant, not only due to the intrinsic importance of mating and self-incompatibility, but due to the fact that isogamy is thought to be the evolutionary precursor to anisogamy: Thus, studying the factors governing the number of mating types in isogamous species may shed light on why anisogamous species have only two distinct types of gametes.

\subsection{Phenotypic clustering and sympatric speciation}
\hl{Write stuff about sympatric speciation, adaptive radiations, and phenotypic clustering.}
\\
\\
In all these cases, we are interested in monitoring the changes in the number of distinct phenotypic entities, in a scenario in which the phenotype is ecologically relevant for fitness. Existing literature has tended to use the word `morph' for this concept, and I will bow down to convention and do the same. However, it is important to remember that in this context, `morphs' are only defined in terms of traits that \textit{affect fitness}, in contrast to the more general use of the word in biology to mean any difference whatsoever (as in single nucleotide polymorphisms, for example). Note that in many of these cases, even though the emergent alternative morphs are discretely distributed, the underlying traits need not be, and indeed are often continuously varying, and only categorized into discrete bins by human experimentalists. Such discrete morphs are often persistent in phylogenetically similar lineages \citep{jamie_persistence_2020}, suggesting they are evolutionarily predictable to some degree, and are also thought to be important for very general evolutionary processes \citep{west-eberhard_alternative_1986}. Intraspecific variation also provides an important metric for the diversity harboured by populations and can have important ecological consequences. Two important questions immediately arise:

\begin{enumerate}
    \item How does a population that is initially monomorphic for a (selectively non-neutral) trait evolve to become polymorphic for this trait in a seemingly discrete manner?
    \item What governs the number of discrete morphs that co-occur in a population?
\end{enumerate}

This very general setting motivates a view in which an `individual' is simply characterized by its phenotype as a point in some abstract `trait space'. These questions then concern the distribution of points in this space (analogous to the frequency of each `type' of individual) changes over evolutionary time due to ecological fitness determinants (through the traits that the individuals possess). More formally, we can visualize an abstract trait space $\mathcal{T}$ which captures all possible phenotypes that an individual may exhibit. Any phenotype can then be viewed as some point $x \in \mathcal{T}$. We are interested in the distribution of trait values in the population subject to some ecological and evolutionary rules, and a `morph' or `eco-variant' is just a cluster within this trait space (a mode in the distribution).
% \subsection{Sympatric speciation}
% The origin of species is relatively well understood when populations diverge in allopatry (separated by geographic barriers, for example). In contrast, sympatric speciation, first formally proposed by Maynard Smith \hl{cite}, is much less well-understood, despite empirical evidence that it has occurred in nature \hl{cite}. Since speciation is one of the fundamental processes in

@@

>>>>

More than 150 years have passed since Ernst Haeckel first coined the term ‘ecology’ in 1866. Today, ecology and evolution are incredibly interdisciplinary, borrowing techniques and ideas from diverse fields such as computer science, statistics, economics, dynamical systems, physics, and information theory. With this development has come a cornucopia of models that use these tools and techniques to try and understand biological phenomena. Though many of these models deal with specific `low-level' factors, there is also value to formulating general, idealized models in abstract terms to underscore these organizing principles and better illustrate the fundamental processes that are required to capture the `essence' of a biological pattern \citep{frank_natural_2012, vellend_theory_2016}. Such abstractions and idealizations are useful in unifying apparently disparate modelling frameworks and are thus vital to theory-building \citep{luque_mirror_2021}.

@@

\section{A brief history of high-level modelling frameworks in population biology}

In population genetics, these high-level forces are the standard evolutionary forces of natural selection, genetic drift, dispersal, and mutation, and the description of evolution in terms of these forces was first laid out in formal mathematical terms during the Modern Synthesis \citep{ewens_mathematical_2004}. The mathematization of the evolutionary forces proved extremely successful, unifying two major schools of thought - Mendelian genetics and Darwinian evolution - that were, at the time, thought to be incompatible \hl{cite something here}.\\
Classical population genetics regarded forces like selection and mutation as fundamental, and the success of this approach during the Modern Synthesis illustrates the value of formulating high-level, abstract models that only provide a `high-level' description of the fundamental processes required to capture the essence of a biological pattern. However, the drama of evolution famously occurs in the ecological theatre \citep{hutchinson_ecological_1965}, and quantities like fitness are not truly fundamental, but instead emerge as the result of various ecological interactions, tradeoffs, and constraints, a fact that can have important consequences for evolution \citep{coulson_putting_2006, kokko_can_2017}. Trying to model such `eco-evolutionary dynamics' has sprouted a rich body of theoretical literature and led to the development of theoretical frameworks like evolutionary game theory and adaptive dynamics, fields which have greatly enriched our understanding of biological populations \citep{brown_why_2016}. Eco-evolutionary population dynamics can broadly be organized under a single unifying framework, the Price equation, that yields all the relevant formal structures under various limits \citep{page_unifying_2002, lion_theoretical_2018}. The Price equation partitions changes in population composition into multiple terms, each of which lends itself to a straightforward interpretation in terms of the high-level evolutionary forces of selection, mutation, and drift, thus providing a useful conceptual framework for thinking about how populations change over time \citep{frank_natural_2012}. The Price equation also leads to a small number of simple yet insightful `fundamental theorems' of population biology \citep{queller_fundamental_2017, lion_theoretical_2018, lehtonen_price_2018} and unifies several various seemingly disjoint formal structures under a single theoretical banner \citep{ lehtonen_price_2020, luque_mirror_2021}\\
The general guiding philosophy of much of this mathematization has been that incorporating the reality of finite population sizes into models leads to no major qualitative differences in behavior, only `adding noise' or `blurring out' the predictions of simpler infinite population models \citep{page_unifying_2002}. Consequently, several major theoretical frameworks in the field, such as adaptive dynamics, are explicitly formulated in deterministic terms at the infinite population size limit. However, this assumption is largely unjustified, and since populations in the real world are finite and stochastic, checking whether stochastic models differ from their deterministic analogs is vital to furthering our understanding of the fundamentals of population biology \citep{hastings_transients_2004, coulson_skeletons_2004, shoemaker_integrating_2020}. Today, we increasingly recognize that incorporating the finite and stochastic nature of the real world routinely has much stronger consequences than simply `adding noise' to deterministic expectations \citep{boettiger_noise_2018} in both ecological \citep{schreiber_does_2022} and evolutionary \citep{delong_stochasticity_2023} models. Stochastic models often exhibit phenomena that do not occur in infinite-population models \citep{rogers_demographic_2012, rogers_spontaneous_2012, veller_drift-induced_2017}, prevent phenomena that occur in infinite populations from occurring \citep{proulx_what_2005, johansson_will_2006, claessen_delayed_2007,  wakano_evolutionary_2013, debarre_evolutionary_2016, johnson_two-dimensional_2021}, or even completely reverse the predictions of deterministic models \citep{houchmandzadeh_selection_2012,houchmandzadeh_fluctuation_2015,constable_demographic_2016,mcleod_social_2019}. Studies of neutral or near-neutral dynamics in population and quantitative genetics usually do take stochasticity seriously, explicitly modeling finite populations that follow stochastic dynamics. However, the classic stochastic models in population and quantitative genetics typically assume a fixed total population size \citep{fisher_genetical_1930, crow_introduction_1970, lande_natural_1976} and their validity is therefore rather restrictive since population sizes routinely fluctuate in the real world. Importantly, the Price equation itself is usually formulated in a deterministic, dynamically-insufficient manner (but see \cite{rice_stochastic_2008} for a stochastic formulation of the Price equation). Since real-life populations are stochastic, finite, and of non-constant population size, it is thus imperative that we develop a theoretical framework that can handle such systems directly, instead of only working with deterministic, infinite-population approximations.\\
Incorporating stochasticity into deterministic systems is a tricky business, and, if done in a phenomenological manner by adding noise to a `deterministic skeleton' \citep{coulson_skeletons_2004} in an ad-hoc fashion, can lead to nonsensical predictions and inconsistencies \citep{strang_how_2019}. Stochastic individual-based models, in which (probabilistic) rules are specified at the level of the individual and population level dynamics are systematically derived from first principles, are much more natural \citep{black_stochastic_2012}, lead to sensible predictions \citep{strang_how_2019}, and can fundamentally differ from the predictions made by simply adding noise terms to a deterministic model \citep{strang_how_2019} (This is also true when incorporating spatial structure into non-spatial models, see for example \cite{durrett_importance_1994}). Formulating the fundamental formal structures of evolutionary biology in terms of the mechanistic demographic processes of birth and death at the individual level is also greatly desirable for biological reasons \citep{metcalf_why_2007,geritz_mathematical_2012}: Since demographic processes such as birth and death rates explicitly account for the ecology of the system, they can more accurately reflect the complex interplay between ecological and evolutionary processes \citep{doebeli_towards_2017} and provide a more fundamental mechanistic description of the relevant evolutionary forces. In other words, `all paths to fitness lead through demography' \citep{metcalf_all_2007}.

\section{Outline of the rest of this thesis}

In this thesis, I present a formulation of population dynamics constructed from mechanistic first principles grounded in individual-level birth and death. Part \ref{part_theory} presents the mathematical formalism in a detailed, self-contained, pedagogical manner. To facilitate readership by a broad audience, I only assume passing familiarity with calculus (derivatives, integrals, Taylor expansions) and probability. Familiarity with stochastic calculus is helpful for some sections but is not required. I present a brief introduction to the relevant mathematics in section \ref{sec_math_background} and present a toy example of tracking population size of a population of identical individuals in section \ref{sec_1D_processes}. I introduce a description of the system via a `master equation', and then conduct a `system-size expansion' to obtain a Fokker-Planck equation for the system. Finally, I conduct a weak noise approximation to arrive at a linear Fokker-Planck equation which can be solved exactly using some stochastic calculus to arrive at a closed-form solution given by a time-dependent Ornstein-Uhlenbeck process, thus illustrating all the major tools required. In section \ref{sec_nD_processes}, I present a multivariate process to describe the evolution of discretely varying traits, and, as before, illustrate the system size expansion and the weak noise approximation. Under mild assumptions, I show that the deterministic limit of this process is the well-known replicator-mutator equation (or equivalently, Eigen's quasispecies equation), thus establishing the microscopic basis of well-known equations in evolutionary game theory. I also show that the mean value of the trait in the population changes according to the Price equation in the deterministic limit. Chapter \ref{chap_infD_processes} introduces a function-valued process to model the evolution of quantitative traits such as body size, which can take on uncountably many values. This function-valued process can then also be analyzed via a system-size approximation to arrive at a `functional' Fokker-Planck equation, in terms of functional derivatives. Under mild assumptions, I show that classic equations such as Kimura's infinite alleles model and the canonical equation of adaptive dynamics can be derived as the deterministic limits of this stochastic process. I also conduct a weak noise approximation to arrive at a linear functional Fokker-Planck equation.  Sections \ref{sec_1D_processes} and \ref{sec_nD_processes} apply techniques that are well-known in physics, as applied to population biology. Chapter \ref{chap_infD_processes} generalizes the work of Tim Rogers and colleagues \citep{rogers_demographic_2012,rogers_spontaneous_2012,rogers_modes_2015}, and to the best of my knowledge, is entirely original. Chapter \ref{chap_infD_processes} also presents a simple heuristic derivation of quantitative genetics models and adaptive dynamics from stochastic first principles that is much simpler than the rigorous mathematical derivations grounded in measure theory and martingale/markov theory that are currently standard reference in theoretical ecology \citep{champagnat_individual_2008}. Examples for one-dimensional, multi-dimensional, and quantitative trait models that can be analyzed in this formalism are presented in Appendix \ref{App_examples}.\\
Part \ref{part_summary} summarizes the major results of this formalism and presents some simple equations that can be argued to be `fundamental theorems' of population biology in the sense of \cite{queller_fundamental_2017}. These theorems reduce to well-known results such as the Price equation, the replicator-mutator equation from evolutionary game theory, and Fisher's fundamental theorem from population genetics in the infinite population limit. For finite populations, these same theorems predict a new evolutionary force, `noise-induced selection', that has still not found its way into the standard formal canon of evolutionary biology such as the Price equation, and whose significance is only recently being recognized \citep{constable_demographic_2016,mcleod_social_2019,mazzolini_universality_2022, kuosmanen_turnover_2022}.

%\section{Coexistence of eco-variants in diverse contexts}\label{synthesis}
%The phenomenon of coexistence of multiple phenotypically distinct, ecologically relevant variants of an organism in sympatry occurs in several different contexts in the natural world. This is best illustrated through examples, which I provide below.
%\subsection{Alternative Reproductive Tactics}
%Alternative reproductive tactics (ARTs) are discrete polymorphisms that occur in sympatry in one or both sexes of a species in response to intrasexual competition for mates. More precisely, two or more traits which are expressed in individuals of a given sex of a given species can be said to be ARTs if \citep{oliveira_alternative_2008}:
%\begin{enumerate}[label=(\alph*)]
%    \item They are \emph{alternative}, in the sense that a given individual can only express one of the behaviors at any given point in time.
%    \item They are \emph{reproductive}, in the sense that the traits are directly relevant to the process of obtaining a mate and are intended to mitigate conspecific intrasexual competition.
%    \item They are \emph{tactics}, in the sense of serving a well-defined adaptive function and thus having fitness consequences. Usually, different ARTs have different associated fitness effects \emph{ceteris paribus}.
%\end{enumerate}
%ARTs are widespread in the animal kingdom, and may be seen either in morphological traits, or as more complex behavioral polymorphisms (see \citep{oliveira_alternative_2008} for an overview). Most (if not all) species which exhibit ARTs only show a small number of polymorphisms (usually 2-3, 5 in a handful of cases, and very rarely 7-8) - \hl{I can maybe make a table for this: Somehow wasn't able to find one in the literature}. ARTs have obvious consequences for reproductive life history, and can strongly influence the ecology and evolutionary trajectory of organisms.
%
%\subsection{Trophic resource polymorphisms}
%
%Trophic resource polymorphism is defined as ``the occurrence of discrete intraspecific morphs showing differential niche use, usually through differences in feeding biology and habitat use" \citep{skulason_resource_1995}. Species which exhibit trophic resource polymorphisms often show marked, discontinuous morphological variations adapted for specific niche usage. For example, arctic charr and african cichlid fishes both often exhibit discrete trophic polymorphisms in which different morphs (of the same species) are specialized for feeding at different strata of the lakes they live in \citep{recknagel_ecosystem_2017}. Such polymorphisms also generally occur in relatively small numbers in sympatry (see Table 1 in \citep{smith_evolutionary_1996}). They are relatively widespread in vertebrates, and are thought to have important evolutionary consequences by acting as the `substrate' for adaptive radiations \citep{smith_evolutionary_1996}. 
%
%\subsection{Mating types in isogamous species}
%Isogamous organisms are sexual, but do not have gametes which can be classified into `male' and `female' based on size. Gametes of such species can, however, be divided into distinct `mating types' according to chemical self-incompatibility. There is substantial variation in the number of mating types present in a given species: While most organisms have only two mating types, this is by no means the rule \citep{phadke_rapid_2009,constable_rate_2018}, with a handful of organisms having 7-10 mating types, and one species of fungus (\textit{Schizophyllum commune}) exhibiting over 23,000 mating types. The study of mating types in isogamous species is ecologically and evolutionarily relevant, not only due to the intrinsic importance of mating and self-incompatibility, but due to the fact that isogamy is thought to be the evolutionary precursor to anisogamy: Thus, studying the factors governing the number of mating types in isogamous species may shed light on why anisogamous species have only two distinct types of gametes.
%
%\subsection{Phenotypic clustering and sympatric speciation}
%\hl{Write stuff about sympatric speciation, adaptive radiations, and phenotypic clustering.}
%\\
%\\
%In all these cases, we are interested in monitoring the changes in the number of distinct phenotypic entities, in a scenario in which the phenotype is ecologically relevant for fitness. Existing literature has tended to use the word `morph' for this concept, and I will bow down to convention and do the same. However, it is important to remember that in this context, `morphs' are only defined in terms of traits that \textit{affect fitness}, in contrast to the more general use of the word in biology to mean any difference whatsoever (as in single nucleotide polymorphisms, for example). Note that in many of these cases, even though the emergent alternative morphs are discretely distributed, the underlying traits need not be, and indeed are often continuously varying, and only categorized into discrete bins by human experimentalists. Such discrete morphs are often persistent in phylogenetically similar lineages \citep{jamie_persistence_2020}, suggesting they are evolutionarily predictable to some degree, and are also thought to be important for very general evolutionary processes \citep{west-eberhard_alternative_1986}. Intraspecific variation also provides an important metric for the diversity harboured by populations and can have important ecological consequences. Two important questions immediately arise:
%
%\begin{enumerate}
%    \item How does a population that is initially monomorphic for a (selectively non-neutral) trait evolve to become polymorphic for this trait in a seemingly discrete manner?
%    \item What governs the number of discrete morphs that co-occur in a population?
%\end{enumerate}
%
%This very general setting motivates a view in which an `individual' is simply characterized by its phenotype as a point in some abstract `trait space'. These questions then concern the distribution of points in this space (analogous to the frequency of each `type' of individual) changes over evolutionary time due to ecological fitness determinants (through the traits that the individuals possess). More formally, we can visualize an abstract trait space $\mathcal{T}$ which captures all possible phenotypes that an individual may exhibit. Any phenotype can then be viewed as some point $x \in \mathcal{T}$. We are interested in the distribution of trait values in the population subject to some ecological and evolutionary rules, and a `morph' or `eco-variant' is just a cluster within this trait space (a mode in the distribution).
% \subsection{Sympatric speciation}
% The origin of species is relatively well understood when populations diverge in allopatry (separated by geographic barriers, for example). In contrast, sympatric speciation, first formally proposed by Maynard Smith \hl{cite}, is much less well-understood, despite empirical evidence that it has occurred in nature \hl{cite}. Since speciation is one of the fundamental processes in

@@

%Raymond Pearl, one of the pioneers of mathematical ecology as a discipline, wrote in 1927 that ``What we want to know is how the biological forces of natality and mortality are so integrated and correlated in their action as to lead to a final result in size of population which follows this particular curve rather than some other one'' \citep{pearl_growth_1976}. Pearl realized 95 years ago that population dynamics must be ultimately explained by the mechanistic processes of birth and death. Today, there are mounting calls for more mechanistic models of evolution that are grounded in these fundamental processes of birth and death \citep{geritz_mathematical_2012,doebeli_towards_2017}. Ecologists also increasingly recognize that incorporating stochasticity is vital to developing more realistic ecological models \citep{hastings_transients_2004, coulson_skeletons_2004, boettiger_noise_2018, shoemaker_integrating_2020,schreiber_does_2022} and does more than `add noise' to deterministic expectations. Individual-based models, where ecological rules are specified at the level of the individual, are a powerful mathematical tool for mechanistic descriptions of stochastic population dynamics \citep{black_stochastic_2012}. Birth-death processes are a very general class of stochastic processes that can be used to capture a wide range of eco-evolutionary processes. `System-size expansions' and their subsequent analysis using the `weak noise approximation' are common tools for analyzing stochastic birth-death processes that are well-known in the statistical physics and applied mathematics communities \citep{gardiner_stochastic_2009}. However, these tools remain relatively underappreciated in ecology, despite being relatively easy to understand and extremely well-motivated in scenarios germane to ecology and evolution. Here, I present a formulation of population dynamics constructed from first principles grounded in birth-death processes. To facilitate readership by a broad audience, only passing familiarity with calculus (derivatives, integrals, Taylor expansions) and probability are assumed. Familiarity with stochastic calculus is helpful for some sections but is not required.  In section \ref{sec_1D_processes}, I show how fluctuating population size of populations of identical individuals can be tracked through a one-dimensional birth-death process. I introduce a description of the system via a `master equation', and then conduct a `system-size expansion' to obtain a Fokker-Planck equation for the system. Finally, I conduct a weak noise approximation to arrive at a linear Fokker-Planck equation which can be solved exactly using some stochastic calculus to arrive at a closed-form solution given by a time-dependent Ornstein-Uhlenbeck process. As an example, I illustrate the complete process for a stochastic version of the logistic equation. In section \ref{sec_nD_processes}, I present a multivariate process to describe the evolution of discretely varying traits, and, as before, illustrate the system size expansion and the weak noise approximation. Under mild assumptions, I show that the deterministic limit of this process is the well-known replicator-mutator equation (or equivalently, Eigen's quasispecies equation), thus establishing the microscopic basis of well-known equations in evolutionary game theory. I also show that the mean value of the trait in the population changes according to the Price equation in the deterministic limit. Chapter \ref{chap_infD_processes} introduces a function-valued process to model the evolution of quantitative traits such as body size, which can take on uncountably many values. This function-valued process can then also be analyzed via a system-size approximation to arrive at a `functional' Fokker-Planck equation, in terms of functional derivatives. Under mild assumptions, I show that classic equations such as Kimura's infinite alleles model and the canonical equation of adaptive dynamics can be derived as the deterministic limits of this stochastic process. I also conduct a weak noise approximation to arrive at a linear functional Fokker-Planck equation.  Sections \ref{sec_1D_processes} and \ref{sec_nD_processes} apply techniques that are well-known in physics, as applied to population biology. Chapter \ref{chap_infD_processes} generalizes the work of Tim Rogers and colleagues \citep{rogers_demographic_2012,rogers_spontaneous_2012,rogers_modes_2015}, and to the best of my knowledge, is entirely original. Chapter \ref{chap_infD_processes} also presents a simple heuristic derivation of quantitative genetics models and adaptive dynamics from stochastic first principles that is much simpler than the rigorous mathematical derivations grounded in measure theory and martingale/markov theory that are currently standard reference in theoretical ecology \citep{champagnat_individual_2008}. I illustrate the utility of all this abstract formalism through examples in Chapter \ref{chap_examples}, and discuss the major implications in \ref{chap_unification}.
